DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-11-05 Paper_May_2012/main.tex
Paper: Vorschlag fuer eine Einleitung und Danksagung

\begin{abstract}
In this article, we describe the software library 
\textit{Differential Equations and Optimization Environment} (\dope{}).
The main feature of \dope{} is that it provides a unified interface to high level algorithms 
such as time stepping methods, nonlinear solvers and optimization routines. This structure ensures 
that first of all the user is required to write those sections of code that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with the need for just a few lines of code to change.
The article illustrates performance and features 
of the software package by four numerical tests. 
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
The \textit{Differential Equations and Optimization Environment} (\dope{}) 
provides a software toolkit to solve forward PDE
problems as well as optimal control problems constrained by a PDE. 
Its main feature is to give a unified interface to high level algorithms such as 
time stepping methods, nonlinear solvers and optimization routines. 
We aim that the user should only need to write those parts
of the code that are problem dependent while all invariant parts of the algorithms
should be reusable without any need for further coding.
In particular, the user should be able to switch between various different 
algorithms without the need to rewrite the problem dependent code, though he or she will
have to replace the algorithm object with an other one.

The
solution of a broad variety of PDE is possible in other software
libraries (like in \deal{} \cite{deal}, dune \cite{dune} , 
or commercial solvers like Ansys \cite{ansys}) 
as well, but
\dope{} concentrates on a unified approach for both linear and nonlinear
problems by interpreting every PDE problem as nonlinear and applying a
Newton method to solve it. While \deal{} leaves much of the work and many
decisions to the user, \dope{} intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimal control problems with PDE
constraints is an innovation in the \dope{} framework.
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and fluid-structure interactions.

The \dope{} project is 
based on the \deal{} \cite{deal} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{deal}.
The authors acknowledge their past experience as well as discussions with 
the authors of the libraries 
Gascoigne/RoDoBo project, which was initiated by 
Roland Becker, Dominik Meidner,  and Boris Vexler \cite{rodobo}. 
From which some of the ideas to modularize the algorithms have arisen.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%The aim of \dope{} is to provide a software toolkit to solve forward PDE
%problems as well as optimal control problems constrained by PDE. The
%solution of a broad variety of PDE is possible in \deal{} as well, but
%\dope{} concentrates on a unified approach for both linear and nonlinear
%problems by interpreting every PDE problem as nonlinear and applying a
%Newton method to solve it. While \deal{} leaves much of the work and many
%decisions to the user, \dope{} intends to be user-optimized by delivering
%prefabricated tools which require from the user only adjustments connected
%to his specific problem. The solution of optimal control problems with PDE
%constraints is an innovation in the \dope{} framework.
%The focus is on the numerical solution of both stationary and nonstationary
%problems which come from different application fields, like elasticity and
%plasticity, fluid dynamics, and fluid-structure interactions.
%

At the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from \deal{} including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Self written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Goal-oriented mesh adaptation with the dual-weighted residual method.
\item Different spatial triangulations for control and state variables.
\end{itemize}

The article is organized as follows. In Section
\ref{detailed_description}

@@

\section*{Thanks}
\todo{Wem wir so danken wollen, Boris, Dominik, Wolfgang ...?}

>>>>

\begin{abstract}
In this article, we describe the software library 
\textit{Differential Equations and Optimization Environment} (\dope{}).
The main feature of \dope{} is that it provides a unified interface to high level algorithms 
such as time stepping methods, nonlinear solvers and optimization routines. This structure ensures 
that first of all the user is required to write those sections of code that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with the need for just a few lines of code to change.
The article illustrates the performance and various features 
of the software package and provides some 
numerical results as demonstration.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exist a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \cite{deal}, dune \cite{dune} , 
or commercial solvers like Ansys \cite{ansys}). 
In fact, these libraries provide very useful tools 
for a huge variety of applications.

However, often, the user is confronted with 
certain bottlenecks. Either, standard high-level routines 
which are necessary to solve even simplest PDEs,
such as linear and nonlinear solvers or timestepping schemes must be implemented 
by the user himself. However, such routines are problem-independent 
and could be provided as standard tools. In fact, in many 
numerical studies, the main task is to compute a  
target functional which represents a certain error norm,
point value, etc.
Second, there is a ever increasing interest in computing 
optimization problems in which the state equation is governed 
by a PDE. Such optimization problems comprise optimal control,
parameter estimation, inverse problems or optimal experimental design. 
So far, there exist a very limited number of open source packages
which provides such routines for scientific studies.

On the other hand, 
commercial toolkits provide these previously mentioned routines but limits user-interaction
with the source code very often. This is quite unsatisfactory because
developement of novel numerical algorithms should enable the programmer
to change certain parameters or pieces of source code.

The novel software toolkit 
\textit{Differential Equations and Optimization Environment} (\dope{}) 
trys to fill this gap. We present a software
to solve forward PDE
problems as well as optimization problems constrained by a PDE. 
Its main feature is to give a unified interface to high level algorithms such as 
time stepping methods, nonlinear solvers and optimization routines. 
We aim that the user should only need to write those parts
of the code that are problem dependent while all invariant parts of the algorithms
should be reusable without any need for further coding.
In particular, the user should be able to switch between various different 
algorithms without the need to rewrite the problem dependent code, though he or she will
have to replace the algorithm object with an other one.

While \deal{} (which might presently seen as the basis because 
it provides various finite elements for discretization) 
leaves much of the work and many
decisions to the user, \dope{} intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimization problems with PDE
constraints is an innovation in the \dope{} framework.
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and fluid-structure interactions.

Summarizing, 
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from \deal{} including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Self written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Goal-oriented mesh adaptation with the dual-weighted residual method.
\item Different spatial triangulations for control and state variables.
\end{itemize}

The article is organized as follows. In the main Section
\ref{detailed_description}, a detailed description of 
the main features is given. In fact, we first explain the solution 
of stationary and nonstationary PDEs and give afterwards some 
ideas how to treat optimization problems. Finally, 
additional features are discussed. In the final Section
\ref{applications}, we discuss three different numerical 
examples which substantiates the features of \dope{}.

@@

\section*{Thanks}
The \dope{} project makes use of various finite elements taken from  
the \deal{} \cite{deal} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{deal}.
The authors acknowledge their past experience as well as discussions 
on modularization of algorithms
with 
the authors of the libraries 
Gascoigne/RoDoBo project, which was initiated by 
Roland Becker, Dominik Meidner, and Boris Vexler \cite{rodobo}. 
The second author thanks Mary F. Wheeler (ICES, Austin) for the 
possibility to finish this work.
